\LARGE \bf Mean Estimation from Adaptive One-bit Measurements

### 2017-09-30 Writeups/onebit_adaptive_Allerton_V3.tex
addressed John's comments in V3, including an extention of Thm. 3 to MSE convergance

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
We consider the problem of estimating the mean of a normal distribution under the following constraint: the estimator can access only a single bit of each sample from this distribution. We study the squared error risk in this estimation as a function of the number of samples and one-bit measurements $n$. We consider an adaptive estimation setting where the single-bit sent at step $n$ is a function of both the new sample and the previous $n-1$ acquired bits. For this setting, we show that no estimator can attain asymptotic mean squared error smaller than $\pi/(2n)$ times the variance. 
%
In other words, one-bit restriction increases the number of samples required for a prescribed accuracy of estimation by a factor of at least $\pi/2$ compared to the unrestricted case.
{\color{red} This fundamental restriction on estimation in the adaptive setting implies a lower bound any distributed estimation setting from one-bit samples. }
%
In addition, we provide an explicit estimator that attains this asymptotic error, showing that, rather surprisingly, only $\pi/2<1.571$ times more samples are required in order to attain estimation performance equivalent to the unrestricted case. 
\end{abstract}

@@

Next, we present an adaptive estimation scheme that attains ARE of $\pi/2$. 
\subsection{Asymptotically optimal estimator}
Consider the following estimator $\widehat{\theta}_n$ for $\theta$:  set 
\begin{equation}
\label{eq:sgd_alg}
\theta_n = \theta_{n-1} +  \gamma_n \sgn (X_n - \theta_n), \quad n = 1,2,\ldots,
\end{equation}
where $\left\{\gamma_n \right\}_{n=1}^\infty$ is any strictly positive sequence satisfying 
\begin{enumerate}
\item[(i)] $\frac{\gamma_n - \gamma_{n+1}}{\gamma_n} = o(\gamma_n)$ \\
\item[(ii)] $\sum_{n=1}^\infty \frac{\gamma_n^{(1+\lambda)/2}} {\sqrt{n}} < \infty$ 
for some $0< \lambda \leq 1$.
\end{enumerate}
(e.g. $\gamma_n = n^{-\beta}$ for $\beta \in (0,1)$). The $n$th step estimation is defined by 
\begin{equation} \label{eq:sgd_est}
\widehat{\theta}_n =  \frac{1}{n} \sum_{i=1}^n  \theta_i. 
\end{equation}

For the estimator defined by \eqref{eq:sgd_est} and \eqref{eq:sgd_alg} 
we have the following results:
\begin{thm} \label{thm:sgd}
The sequence $\widehat{\theta}_n$ of \eqref{eq:sgd_est} satisfies
\[
\sqrt{n} \left( \widehat{\theta}_n - \theta \right) \overset{d}{\rightarrow} \mathcal N \left(0,  \pi \sigma^2 /2 \right).
\]
\end{thm}

\subsubsection*{Proof}
The asymptotic behavior of \eqref{eq:sgd_est} is a special case of  \cite[Thm. 4]{polyak1992acceleration}. The details are provided in the Appendix.\\

@@

\subsection*{Proof of Thm.~\ref{thm:sgd}}
The algorithm given in \eqref{eq:sgd_alg} and \eqref{eq:sgd_est} is a special case of a more general class of estimation procedures given in \cite{polyak1992acceleration}. In particular, Thm.~\ref{thm:sgd} is follows directly from the following simplified version of \cite[Thm. 4]{polyak1992acceleration}:
\begin{thm}{\cite[Thm. 4]{polyak1992acceleration}} \label{thm:polyak_juditsky}
Let 
\[
X_i = \theta + Z_i,\quad i=1,\ldots,n,
\]
where the $Z_i$s are i.i.d. with zero mean. Define 
\begin{align*}
\theta_i & = \theta_{i-1} + \gamma_i \varphi(X_i - \theta_{i-1}), \\
\widehat{\theta}_n & = \frac{1}{n} \sum_{i=0}^{n-1} \theta_i,
\end{align*}
where the sequence $\left\{ \gamma_i \right\}_{i=1}^\infty$ satisfies conditions (i) and (ii) in Thm.~\ref{thm:sgd}, and $\left| \varphi(x) \right| \leq K_1(1+x)$ for some $K_1$. Define $\psi(x) = \mathbb E \varphi(x+Z_1)$, $\chi(x) = \mathbb E \varphi^2(x+Z_1)$ and assume that $\psi(0)=0$, $x\psi(x) >0$ for all $x\neq 0$, $\chi(x)$ is continuous at zero, and that $\psi(x)$ is differentiable at zero with $\psi'(0)>0$. Moreover, assume that there exists $K_2$ and $0<\lambda \leq 1$ such that
\[
\left| \psi(x) - \psi'(0)x \right|\leq K_2 |x|^{1+\gamma}.
\]
Then $\widehat{\theta}_n \rightarrow \theta$ almost surely and $ \sqrt{n}(\widehat{\theta}_n - \theta)$ converges in distribution to $\mathcal N(0,V)$, where
\[
V = \frac{ \chi(0)} {\psi'^2(0)}. 
\]
\end{thm}

Using the notation in Thm.~\ref{thm:polyak_juditsky}, we set $\varphi(x) = \sgn(x)$ and $Z_i = X_i - \theta$. We have: 
\[
\chi(x) = \mathbb E \varphi^2(x+Z_1),
\]
so $\chi(0) = 1$. In addition,
\begin{align*}
\psi(x) & = \mathbb E \sgn(x+ Z_1) = \int_{-\infty}^\infty \sgn(x+z) \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{z^2}{2\sigma^2}} dz \\
& = \int_{-x}^\infty \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{z^2}{2\sigma^2}} dz -\int_{-\infty}^{-x} \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{z^2}{2\sigma^2}} dz.
\end{align*}
This leads to 
\begin{align*}
\psi'(x) & = \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{x^2}{2\sigma^2}} dz +\frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{z^2}{2\sigma^2}} dz,
\end{align*}
so $\psi'(0) = \frac{2}{\sqrt{2\pi}\sigma}$. It is now easy to verify that the rest of the conditions in Thm.~\ref{thm:polyak_juditsky} are fulfilled for any $\lambda > 0$. Since 
\[
\frac{\chi(0)}{\psi'^2(0)} = \frac{\pi \sigma^2}{2},
\]
Thm.~\ref{thm:sgd} follows from Thm.~\ref{thm:polyak_juditsky}. \QEDA

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
We consider the problem of estimating the mean of a normal distribution under the following constraint: the estimator can access only a single bit of each sample from this distribution. We study the squared error risk in this estimation as a function of the number of samples and one-bit measurements $n$. We consider an adaptive estimation setting where the single-bit sent at step $n$ is a function of both the new sample and the previous $n-1$ acquired bits. For this setting, we show that no estimator can attain asymptotic mean squared error smaller than $\pi/(2n)$ times the variance. 
%
In other words, one-bit restriction increases the number of samples required for a prescribed accuracy of estimation by a factor of at least $\pi/2$ compared to the unrestricted case.
%{\color{red} This fundamental restriction on estimation in the adaptive setting implies a lower bound any distributed estimation setting from one-bit samples. }
%
In addition, we provide an explicit estimator that attains this asymptotic error, showing that, rather surprisingly, only $\pi/2<1.571$ times more samples are required in order to attain estimation performance equivalent to the unrestricted case. 
\end{abstract}

@@

Next, we present an adaptive estimation scheme that attains ARE of $\pi/2$. 
\subsection{Asymptotically optimal estimator}

Let $\left\{\gamma_n \right\}_{n=1}^\infty$ be a strictly positive sequence satisfying:
\begin{equation} \label{eq:conditions}
\begin{cases}
\frac{\gamma_n - \gamma_{n+1}}{\gamma_n} = o(\gamma_n), &  \\
\sum_{n=1}^\infty \frac{\gamma_n^{(1+\lambda)/2}} {\sqrt{n}} < \infty, & 
\mathrm{for~some~}0< \lambda \leq 1
\end{cases}
\end{equation}
(e.g. $\gamma_n = n^{-\beta}$ for $\beta \in (0,1)$). Consider the following estimator $\widehat{\theta}_n$ for $\theta$:  
\begin{equation}
\label{eq:sgd_alg}
\theta_n = \theta_{n-1} +  \gamma_n \sgn (X_n - \theta_{n-1}), \quad n = 1,2,\ldots,
\end{equation}
and set the $n$th step estimation as
\begin{equation} \label{eq:sgd_est}
\widehat{\theta}_n =  \frac{1}{n} \sum_{i=1}^n  \theta_i. 
\end{equation}

%For the estimator defined by \eqref{eq:sgd_est} and \eqref{eq:sgd_alg} we have the following results:
\begin{thm} \label{thm:sgd}
The sequence $\widehat{\theta}_n$ of \eqref{eq:sgd_est} satisfies
\begin{enumerate}
\item[(i)]
\[
\sqrt{n} \left( \widehat{\theta}_n - \theta \right) \overset{d}{\rightarrow} \mathcal N \left(0,  \pi \sigma^2 /2 \right).
\]
\item[(ii)] Assume moreover that $\gamma_n = o(n^{-2/3})$ and $\sum_{n=1}^\infty \gamma_n = \infty$. Then
\[
\lim_{n\rightarrow \infty} n\mathbb E \left[ \left(\theta-\widehat{\theta}_n \right)^2 \right] = \sigma^2 \frac{\pi}{2}. 
\]
\end{enumerate}

\end{thm}

\subsubsection*{Proof}
The asymptotic behavior of \eqref{eq:sgd_est} is a special case of \cite[Thm. 4]{polyak1992acceleration} and \cite[Thm. 2]{polyak1990new}. The details are provided in the Appendix.\\

@@

\subsection*{Proof of Thm.~\ref{thm:sgd}}
The algorithm given in \eqref{eq:sgd_alg} and \eqref{eq:sgd_est} is a special case of a more general class of estimation procedures given in \cite{polyak1992acceleration} and \cite{polyak1990new}. Specifically, (i) in Thm.~\ref{thm:sgd} follows directly from the following simplified version of \cite[Thm. 4]{polyak1992acceleration}:
\begin{thm}{\cite[Thm. 4]{polyak1992acceleration}} \label{thm:polyak_juditsky}
Let 
\[
X_i = \theta + Z_i,\quad i=1,\ldots,n,
\]
where the $Z_i$s are i.i.d. with zero means and finite variances. Define
\begin{align*}
\theta_i & = \theta_{i-1} + \gamma_i \varphi(X_i - \theta_{i-1}), \\
\widehat{\theta}_n & = \frac{1}{n} \sum_{i=0}^{n-1} \theta_i, 
\end{align*}
where in addition, assume the following: 
\begin{enumerate}
\item[(i)] There exits $K_1$ such that $\left| \varphi(x) \right| \leq K_1(1+|x|)$ for all $x\in \in \mathbb R$.
\item[(ii)] The sequence $\left\{ \gamma_i \right\}_{i=1}^\infty$ satisfies conditions \eqref{eq:conditions}.
\item[(iii)] The function $\psi(x) \triangleq \mathbb E \varphi(x+Z_1)$ is differentiable at zero with $\psi'(0)>0$, and satisfies $\psi(0)=0$ and $x\psi(x) >0$ for all $x\neq 0$.
Moreover, assume that there exists $K_2$ and $0<\lambda \leq 1$ such that
\begin{equation}
\label{eq:Polyak_Juditsky_cond3}
\left| \psi(x) - \psi'(0)x \right|\leq K_2 |x|^{1+\lambda}.
\end{equation}
\item[(iv)] The function 
$\chi(x) \triangleq \mathbb E \varphi^2(x+Z_1)$ is continuous at zero. 
\end{enumerate}
Then $\widehat{\theta}_n \rightarrow \theta$ almost surely and $ \sqrt{n}(\widehat{\theta}_n - \theta)$ converges in distribution to $\mathcal N(0,V)$, where
\[
V = \frac{ \chi(0)} {\psi'^2(0)}. 
\]
\end{thm}

Using the notation above, we set $\varphi(x) = \sgn(x)$ and $Z_i = X_i - \theta$. We have that $\chi(x) = \mathbb E \sgn^2(x+Z_1) = 1$, so $\chi(0) = 1$. In addition,
\begin{align*}
\psi(x) & = \mathbb E \sgn(x+ Z_1) = \int_{-\infty}^\infty \sgn(x+z) \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{z^2}{2\sigma^2}} dz \\
& = \int_{-x}^\infty \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{z^2}{2\sigma^2}} dz -\int_{-\infty}^{-x} \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{z^2}{2\sigma^2}} dz.
\end{align*}
This leads to 
\begin{align*}
\psi'(x) & = \frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{x^2}{2\sigma^2}} dz +\frac{1}{\sqrt{2\pi}\sigma} e^{-\frac{z^2}{2\sigma^2}} dz,
\end{align*}
so $\psi'(0) = \frac{2}{\sqrt{2\pi}\sigma}$. It is now easy to verify that the rest of the conditions in Thm.~\ref{thm:polyak_juditsky} are fulfilled for any $\lambda > 0$. Since 
\[
\frac{\chi(0)}{\psi'^2(0)} = \frac{\pi \sigma^2}{2},
\]
Thm.~\ref{thm:sgd}-(i) follows from Thm.~\ref{thm:polyak_juditsky}.  \\
In order to prove Thm.~\ref{thm:sgd}-(ii), we consider:
\begin{thm}{ \cite[Thm. 2]{polyak1990new}} \label{thm:polyak_new}
Let
\begin{align} \label{eq:polyak_new_measurements}
\begin{cases}
U_n = U_{n-1} - \gamma_n \varphi(Y_n), & Y_n = f'(U_{n-1})+Z_n \\
\bar{U}_n= \frac{1}{n} \sum_{i=1}^n U_n, & n=1,2,\ldots.
\end{cases}
\end{align}
Assume that the function $f(x)$ is twice differentiable with a strictly positive and uniformly bounded second derivative. In particular, $f(x)$ is convex with a unique minimizer $x^\star \in \mathbb R$. Moreover, assume that the noises $Z_n$ are uncorrelated and identically distributed with a distribution for which the Fisher information exits. Let $\psi(x)$ and $\chi(x)$ be defined as in  Thm.~\ref{thm:polyak_juditsky}-(iii) and satisfies the conditions there. Assume in addition that $\chi(0)>0$, condition \eqref{eq:Polyak_Juditsky_cond3} with $\lambda = 1$, 
and there exits $K_3$ such that 
\[
\mathbb E \left[ | \varphi(x+Z_1) |^4 \right] \leq K_3(1+|x|^4). 
\]
Finally, assume that the sequence $\{\gamma_n \}$ satisfies conditions \eqref{eq:conditions} and the additional conditions in Thm.~\ref{thm:sgd}-(ii). Then
\[
V_n \triangleq \mathbb E \left[ \left(\bar{U}_n-x^\star \right)^2 \right] = n^{-1}\frac{\chi(0)} { (\psi'(0))^2 (f''(x^\star))^2 } + o(n^{-1}).
\]
\end{thm}

We now use Thm.~\ref{thm:polyak_new} with $f(x) = 0.5(\theta-x)^2$, $\varphi(x) = -\sgn(-x)$, $Z_n = -Z_n$ and $U_n = \theta_n$. From \eqref{eq:polyak_new_measurements} we have
\begin{align*} 
\theta_n & = \theta_{n-1} + \gamma_n \sgn(\theta-\theta_{n-1} - Z_n )  \\
& = \theta_{n-1} + \gamma_n \sgn(X_n-\theta_{n-1} ),
\end{align*}
so the estimator $\widehat{\theta}_n$ defined by $\widehat{\theta}_n$ equals to the one defined by \eqref{eq:sgd_est} and \eqref{eq:sgd_alg}. Note that
\[
\mathbb E \left[ | \varphi(x+Z_1) |^4 \right] = 1 \leq K_3(1+|x|^4)
\]
for any $K_3\geq 1$, the Fisher information of $Z_1$ is $\sigma^2$, $\chi(x) = 1 > 0$, and that 
the conditions in Thm.~\ref{thm:polyak_new} on $\psi(x)$ and $\chi(x)$ were verified to hold in the first part of the proof. In particular, $\psi'(0) = \sqrt{2}/ (\sqrt{\pi}/\sigma)$. Since $f(x)$ satisfies the conditions above with $x^\star = \theta$ and $f''(x) = 1$. Thm.~\ref{thm:polyak_new} implies 
\[
n V_n = \mathbb E \left[ \left(\widehat{\theta}_n-\theta \right)^2 \right]  = \frac{\pi}{2}  + o(1).
\]
